Probability

### 2016-04-29 prob.tex
interval

\makeatletter
\newtheoremstyle{indented}
  {3pt}% space before
  {3pt}% space after
  {\addtolength{\@totalleftmargin}{3.5em}
   \addtolength{\linewidth}{-3.5em}
   \parshape 1 3.5em \linewidth}% body font
  {}% indent
  {\bfseries}% header font
  {.}% punctuation
  {.5em}% after theorem header
  {}% header specification (empty for default)
\makeatother
\theoremstyle{indented}
%% \theoremstyle{remark}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newtheorem{cor}{Corollary}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\qevents}{\ensuremath{\mathcal{E}}}
\newcommand{\pmeas}{\ensuremath{\mu}}
\newcommand{\Hilb}{\mathcal{H}}
\newcommand{\pr}[2]{\langle #1, #2 \rangle}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\bra}[1]{\langle#1|}
\newcommand{\ip}[2]{\langle #1 | #2 \rangle}
\newcommand{\proj}[1]{|#1 \rangle\langle #1 |}
\newcommand{\ps}{\texttt{+}}
\newcommand{\ms}{\texttt{-}}
\newcommand{\poss}{{\mbox{\wesa{possible}}}}
\newcommand{\imposs}{{\mbox{\wesa{impossible}}}}
\newcommand{\likely}{{\mbox{\wesa{likely}}}}
\newcommand{\unlikely}{{\mbox{\wesa{unlikely}}}}
\newcommand{\overflow}{{\mbox{\wesa{overflow}}}}

@@

\begin{example}[Two-coin probability space]
  Consider an experiment that tosses two coins. We have four possible
  outcomes that constitute the sample space
  $\Omega = \{ HH, HT, TH, TT \}$. There are 16 total events including
  for example the event $\{ HH, HT \}$ that the first coin is
  ``heads,'' the event $\{ HT, TH \}$ that the two coins land on
  opposite sides, and the event $\{ HT, TH, TT\}$ that at least one
  coin is tails. Here is a possible probability measure for these
  events:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) &=& 0 \\
\pmeas(\{ HH \}) &=& 1/3 \\
\pmeas(\{ HT \}) &=& 0 \\
\pmeas(\{ TH \}) &=& 2/3 \\
\pmeas(\{ TT \}) &=& 0 \\
\pmeas(\{  HH, HT \}) &=& 1/3 \\
\pmeas(\{  HH, TH \}) &=& 1 \\
\pmeas(\{  HH , TT \}) &=& 1/3 
\end{array} & \begin{array}{rcl}
\pmeas(\{  HT, TH \}) &=& 2/3 \\
\pmeas(\{  HT , TT \}) &=& 0 \\
\pmeas(\{  TH , TT \}) &=& 2/3 \\
\pmeas(\{  HH, HT, TH \}) &=& 1 \\
\pmeas(\{  HH, HT, TT \}) &=& 1/3 \\
\pmeas(\{  HH, TH, TT \}) &=& 1 \\
\pmeas(\{  HT, TH, TT \}) &=& 2/3 \\
\pmeas(\{  HH, HT, TH, TT \}) &=& 1
\end{array}
\end{array}\]
The assignment satisfies the two constraints for probability measures:
the probability of the entire sample space is 1, and the probability
of every collection of disjoint events (e.g.,
$\{ HT, TH \} = \{ HT \} \cup \{ TH \}$) is the sum of the individual
probabilities. The probability of collections of non-disjoint events
(e.g., $\{ HT, TH, TT \} = \{ HT, TH \} \cup \{ TH , TT \}$) may add
to something different than the probabilities of the individual
events. It is useful to think that this probability measure is
completely induced by the two coins in question and their
characteristics in the sense that each pair of coins induces a
measure, and each measure must correspond to some pair of coins.
In this example, the probability measure is induced by a weighted first coin and
a double-headed second coin.
\qed\end{example}

In a strict computational or experimental setting, one may question
the reliance of the definition of probability space on the uncountable
and uncomputable real interval $[0,1]$. This interval includes numbers
like $0.h_1h_2h_3\ldots$ where $h_i$ is 1 or 0 depending on whether
Turing machine $M_i$ halts or not. Such numbers cannot be
computed. This interval also includes numbers like $\frac{\pi}{4}$
which can only be computed with increasingly large resources as the
precision increases.
% \yutsung{Check the meaning of the ``computing $\frac{\pi}{4}$'', because
% Bailey Borwein Plouffe formula can computing the $n$th binary digit of $\pi$ 
% using base 16 directly.}
Therefore, in a resource-aware setting, it is more appropriate to
consider probability measures that map events to a finite set of
elements computable with a fixed set of
resources~\cite{Artstein1972,PuriRalescu1983}. The simplest such set,
and the one we will consider exclusively in this paper, is the set
$\mathscr{L}_{2}=\left\{ \imposs, \poss\right\}$ together with the
operation $\vee$ where $x \vee y=\imposs$ if and only if
$x=y=\imposs$. In relation to the first definition, one can interpret
\imposs\ as the closed interval $[0,0]$, \poss\ as the half-open
interval $(0,1]$, and $\vee$ as the addition of intervals. The
definition of a probability measure in this case is modified as being
a function $\pmeas : \events \rightarrow \mathscr{L}_{2}$ such that:
\begin{itemize}
\item $\pmeas(\Omega) = \poss$, and
\item for a collection $E_i$,  of pairwise disjoint events, $\pmeas(\bigcup_i
E_i) = \bigvee_i \pmeas(E_i)$.
\end{itemize}

\begin{example}[Two-coin probability space with finite set-valued
  probability measure] Under the new set-valued requirement, the
  probability measure in the first example becomes:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) &=& \imposs \\
\pmeas(\{ HH \}) &=& \poss \\
\pmeas(\{ HT \}) &=& \imposs \\
\pmeas(\{ TH \}) &=& \poss \\
\pmeas(\{ TT \}) &=& \imposs \\
\pmeas(\{  HH, HT \}) &=& \poss \\
\pmeas(\{  HH, TH \}) &=& \poss \\
\pmeas(\{  HH , TT \}) &=& \poss \\
\end{array} & \begin{array}{rcl}
\pmeas(\{  HT, TH \}) &=& \poss \\
\pmeas(\{  HT , TT \}) &=& \imposs \\
\pmeas(\{  TH , TT \}) &=& \poss \\
\pmeas(\{  HH, HT, TH \}) &=& \poss \\
\pmeas(\{  HH, HT, TT \}) &=& \poss \\
\pmeas(\{  HH, TH, TT \}) &=& \poss \\
\pmeas(\{  HT, TH, TT \}) &=& \poss \\
\pmeas(\{  HH, HT, TH, TT \}) &=& \poss 
\end{array}
\end{array}\]
Compare to the first example, we can still get the information that the second
coin is double-headed, although we loss the information that the first coin is
weighted.
\qed\end{example}

@@

\caption{\label{tab:The-operator-on}The operator~$\vee$ on $\mathscr{L}_{5}=\left\{ \imposs,\poss,\unlikely,\likely,\overflow\right\} $}
\end{table}
\begin{example}{[}Two-coin probability space with $\mathscr{L}_{5}$-valued
probability measure{]} Under the new $\mathscr{L}_{5}$-valued requirement,
the probability measure in the first example becomes: 
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \likely\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \likely\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \likely\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \likely
\end{array}\end{array}
\]
In this example, we can get the information that the first coin is
weighted and the second coin is double-headed. \qed\end{example}

>>>>

\makeatletter
\newtheoremstyle{indented}
  {3pt}% space before
  {3pt}% space after
  {\addtolength{\@totalleftmargin}{3.5em}
   \addtolength{\linewidth}{-3.5em}
   \parshape 1 3.5em \linewidth}% body font
  {}% indent
  {\bfseries}% header font
  {.}% punctuation
  {.5em}% after theorem header
  {}% header specification (empty for default)
\makeatother
\theoremstyle{indented}
%% \theoremstyle{remark}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newtheorem{cor}{Corollary}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\qevents}{\ensuremath{\mathcal{E}}}
\newcommand{\pmeas}{\ensuremath{\mu}}
\newcommand{\Hilb}{\mathcal{H}}
\newcommand{\pr}[2]{\langle #1, #2 \rangle}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\bra}[1]{\langle#1|}
\newcommand{\ip}[2]{\langle #1 | #2 \rangle}
\newcommand{\proj}[1]{|#1 \rangle\langle #1 |}
\newcommand{\ps}{\texttt{+}}
\newcommand{\ms}{\texttt{-}}
\newcommand{\poss}{{\mbox{\wesa{possible}}}}
\newcommand{\imposs}{{\mbox{\wesa{impossible}}}}
\newcommand{\likely}{{\mbox{\wesa{likely}}}}
\newcommand{\unlikely}{{\mbox{\wesa{unlikely}}}}
\newcommand{\necess}{{\mbox{\wesa{necessary}}}}
\newcommand{\overflow}{{\mbox{\wesa{overflow}}}}

@@

\begin{example}[Two-coin probability space]
  Consider an experiment that tosses two coins. We have four possible
  outcomes that constitute the sample space
  $\Omega = \{ HH, HT, TH, TT \}$. There are 16 total events including
  for example the event $\{ HH, HT \}$ that the first coin is
  ``heads,'' the event $\{ HT, TH \}$ that the two coins land on
  opposite sides, and the event $\{ HT, TH, TT\}$ that at least one
  coin is tails. Here is a possible probability measure for these
  events:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) &=& 0 \\
\pmeas(\{ HH \}) &=& 1/3 \\
\pmeas(\{ HT \}) &=& 0 \\
\pmeas(\{ TH \}) &=& 2/3 \\
\pmeas(\{ TT \}) &=& 0 \\
\pmeas(\{  HH, HT \}) &=& 1/3 \\
\pmeas(\{  HH, TH \}) &=& 1 \\
\pmeas(\{  HH , TT \}) &=& 1/3 
\end{array} & \begin{array}{rcl}
\pmeas(\{  HT, TH \}) &=& 2/3 \\
\pmeas(\{  HT , TT \}) &=& 0 \\
\pmeas(\{  TH , TT \}) &=& 2/3 \\
\pmeas(\{  HH, HT, TH \}) &=& 1 \\
\pmeas(\{  HH, HT, TT \}) &=& 1/3 \\
\pmeas(\{  HH, TH, TT \}) &=& 1 \\
\pmeas(\{  HT, TH, TT \}) &=& 2/3 \\
\pmeas(\{  HH, HT, TH, TT \}) &=& 1
\end{array}
\end{array}\]
The assignment satisfies the two constraints for probability measures:
the probability of the entire sample space is 1, and the probability
of every collection of disjoint events (e.g.,
$\{ HT, TH \} = \{ HT \} \cup \{ TH \}$) is the sum of the individual
probabilities. The probability of collections of non-disjoint events
(e.g., $\{ HT, TH, TT \} = \{ HT, TH \} \cup \{ TH , TT \}$) may add
to something different than the probabilities of the individual
events. It is useful to think that this probability measure is
completely induced by the two coins in question and their
characteristics in the sense that each pair of coins induces a
measure, and each measure must correspond to some pair of coins.  In
this example, the first coin is twice as likely to be tails than heads
and the second coin is double-headed.  
\qed\end{example}

In a strict computational or experimental setting, one may question
the reliance of the definition of probability space on the uncountable
and uncomputable real interval $[0,1]$. This interval includes numbers
like $0.h_1h_2h_3\ldots$ where $h_i$ is 1 or 0 depending on whether
Turing machine $M_i$ halts or not. Such numbers cannot be
computed. This interval also includes numbers like $\frac{\pi}{4}$
which can only be computed with increasingly large resources as the
precision increases.
% \yutsung{Check the meaning of the ``computing $\frac{\pi}{4}$'', because
% Bailey Borwein Plouffe formula can computing the $n$th binary digit of $\pi$ 
% using base 16 directly.}
Therefore, in a resource-aware setting, it is more appropriate to
consider probability measures that map events to a finite set of
elements computable with a fixed set of resources. We will consider
two approaches: set-valued probability
measures~\cite{Artstein1972,PuriRalescu1983} and interval-valued
probability measures~\cite{JamisonLodwick2004,Weichselberger2000}.

\paragraph*{Set-valued Probability Measures.} The simplest such set,
and the one we will consider exclusively in this paper, is the set
$\mathscr{L}_{2}=\left\{ \imposs, \poss\right\}$ together with the
operation $\vee$ where $x \vee y=\imposs$ if and only if
$x=y=\imposs$. In relation to the conventional definition, one can
interpret \imposs\ as the closed interval $[0,0]$, \poss\ as the
half-open interval $(0,1]$, and $\vee$ as the addition of
intervals. The definition of a probability measure in this case is
modified as being a function
$\pmeas : \events \rightarrow \mathscr{L}_{2}$ such that:
\begin{itemize}
\item $\pmeas(\Omega) = \poss$, and
\item for a collection $E_i$,  of pairwise disjoint events, $\pmeas(\bigcup_i
E_i) = \bigvee_i \pmeas(E_i)$.
\end{itemize}

\begin{example}[Two-coin probability space with finite set-valued
  probability measure] Under the new set-valued requirement, the
  probability measure in the first example becomes:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) &=& \imposs \\
\pmeas(\{ HH \}) &=& \poss \\
\pmeas(\{ HT \}) &=& \imposs \\
\pmeas(\{ TH \}) &=& \poss \\
\pmeas(\{ TT \}) &=& \imposs \\
\pmeas(\{  HH, HT \}) &=& \poss \\
\pmeas(\{  HH, TH \}) &=& \poss \\
\pmeas(\{  HH , TT \}) &=& \poss \\
\end{array} & \begin{array}{rcl}
\pmeas(\{  HT, TH \}) &=& \poss \\
\pmeas(\{  HT , TT \}) &=& \imposs \\
\pmeas(\{  TH , TT \}) &=& \poss \\
\pmeas(\{  HH, HT, TH \}) &=& \poss \\
\pmeas(\{  HH, HT, TT \}) &=& \poss \\
\pmeas(\{  HH, TH, TT \}) &=& \poss \\
\pmeas(\{  HT, TH, TT \}) &=& \poss \\
\pmeas(\{  HH, HT, TH, TT \}) &=& \poss 
\end{array}
\end{array}\]
Despite the fact that we lost all numeric information, the probability
measure still reveals that the second coin is double-headed. We have
however lost the information regarding the bias in the first
coin. This information can be recovered with a more refined
probability measure as we show next. 
\qed\end{example}

\paragraph*{Interval-valued probability measures.} In this case, we
split the $[0,1]$ interval into a collection of closed sub-intervals
obeying certain restrictions to be described at the end of the
section. To illustrate the main ideas, we first consider a simple
example.

\begin{example}[Two-coin probability space with four intervals]
We split the interval in the following four closed sub-intervals:
\begin{itemize}
\item $[0,0]$ which we call \imposs, 
\item $[0,\frac{1}{2}]$ which we call \unlikely, 
\item $[\frac{1}{2},1]$ which we call \likely, and 
\item $[1,1]$ which we call \necess.
\end{itemize}
The probability measure in this case becomes:
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \necess\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \necess\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \necess\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \necess
\end{array}\end{array}
\]
Not only does this probability measure reveal that the second coin is
double-headed but it also reveals that the first coin is more likely
to be heads than tails, despite the absence of numeric information.
\qed\end{example}

\newpage

\[\begin{array}{rclrcll}
\imposs \vee x &=& x \vee \imposs &=& x \\
\necess \vee x &=& x \vee \necess  &=& \necess \\
\unlikely \vee \likely &=& \likely \vee \unlikely &=& \necess
\end{array}\]

@@

\caption{\label{tab:The-operator-on}The operator~$\vee$ on $\mathscr{L}_{5}=\left\{ \imposs,\poss,\unlikely,\likely,\overflow\right\} $}
\end{table}
